# All Raman Spectra -> Pirouette
## User Guide
This R report generates an input file for pirouette using all raman spectra with a particular *SubjectID*. This is 
to be used to load and visualize all spectra in pirouette at once, but it does not match to HPLC data and cannot be 
used to build a model.
   
### Required Libraries
* Rlabkey
* glue
   
### Adjustable Parameters
The adjustable paramenters are documented throughout the report, but the most essential are directly
below:
    
* *raman_table* corresponds to the name of the dataset containing the raman in this study
* *subject_id* corresponds to the SubjectID you want to generate a file from.

```{r}
raman_table = 'Raman_Transformation'
subject_id = 'B30_35_ATCC32359_WT_1'
```

```{r, echo=FALSE}
library(Rlabkey)
library(glue)

# Verbose
options(echo=TRUE)

base_url = labkey.url.base
folder_path = labkey.url.path
```

# Querying 

## Query Commands

The sql strings are first constructed using glue. Any value inside {}
is interpreted as a variable and is replaced in the returend string with the
corresponding value. This allows changing of dataset names.

Ensure that the column names correspond to *Field Names* not *Caption Names.*
    
### Raman Query
All intensities of the specified SubjectID are pulled. Change the desired SubjectID at
the top of the script.

```{r}
sql_raman = glue("SELECT DISTINCT {raman_table}.SubjectID,
                 {raman_table}.FileName,
                 {raman_table}.Intensities,
				 {raman_table}.Run.ProcessID	
                 FROM {raman_table}
                 WHERE {raman_table}.SubjectID.SubjectID = '{subject_id}'")
```

### Wavenumbers Query
This query gets the wavenumbers corresponding to the SubjectID defined at the top of the script.

The raman dataset is **required** to have a *Run* property called *Wavenumbers* that is a LookUp to 
a list. That list must have the column *WaveNumbers*, which holds the wavenumbers for that configuration.
The list is not required to be in the current study if the raman dataset was imported from an assay, as the
LookUp will automatically reference the orginal list in the assay.
   
```{r}
# Getting wavenumbers from raman table -- should only be one when generating piroutte file.
sql_wavenumbers = glue("SELECT DISTINCT
                       {raman_table}.Run.Wavenumbers.WaveNumbers
                       FROM {raman_table}
                       WHERE {raman_table}.SubjectID.SubjectID = '{subject_id}'")

```
   
## Execution of Queries

This section executes the queries defined above and stores them in a dataframe.

The datasets are assumed to be in a *study* and in the **current folder**.
These can be changed by adjusting the *folderPath* parameter (currently set to current folder by LabKey)
and the *schemaName* parameter (for example, from 'study' to 'assay').

The *colNameOpt* parameter tells the query use Field Names in the data frame, as opposed to caption names.

Finally, the data frames returned are copied because LabKey protects them internally, so this is required to use the
data.table format below.
   
```{r}
df_raman = labkey.executeSql(baseUrl = base_url, folderPath = folder_path, schemaName='study',
                                  sql = sql_raman, colNameOpt ='fieldname')

df_wavenum = labkey.executeSql(baseUrl = base_url, folderPath = folder_path, schemaName='study',
                                    sql = sql_wavenumbers, colNameOpt = 'fieldname')

```
## Writing the Piroutte File

This section is responsible for constructing the piroutte file, which is then made avialable to download
for the user by LabKey.

The order is described here:
   
1. The number of intensities are found, which is needed for the dimension line of the input. This is done
by counting the commas in an wavenumber entry (because they are stored as one continuous string) and adding one.

2. The first line of the file is written, which is the dimension line. It is of the format (# of intensities x # of rows)
   
3. The column line is written, which contains the wavenumbers

4. The rows of data are looped through and each written. The *FileName* value is used as the *ElapsedTime*_*ProcessID*, which piroutte requires. Each row contains the intensities for that file.
   
**Note:** Anytime that you write to a file in LabKey, you must include the comment **# ${fileout:tsvfile** *(closing bracket needed but not shown)* directly above the line with the write call, for **every** write call. *fileout* specifies that you want to make a file
available to download and *tsvfile* is the name of it that you will refer to in the write call. For example:  
   
   # ${fileout:tsvfile  
   write(dimension_line, "tsvfile")
   
See source code for clarification, as LabKey makes the subsitutes in the report.
                                
```{r}

# Get how many intensity entries there are, required for the input dimension size.
# There are stored as a string, so we count the occurances of commas and add one
intensities_len = lengths(regmatches(df_wavenum[1,1], gregexpr(",", df_wavenum[1,1]))) + 1


# First line of file, dimension (number of intensities and rows of data)
dimension_line = paste("#D", paste(intensities_len, nrow(df_raman), sep='x'), sep=', ')


# EVERYTIME you want to WRITE TO A FILE, the following comment is REQUIRED on the line directly above it
# due to how the LabKey variable subsitutor works

# ${fileout:tsvfile}
write(dimension_line, "tsvfile")

# Write the column names (wavenumber and other variable names)
column_line = paste('#C', df_wavenum[1,1], sep = ', ')
# ${fileout:tsvfile}
write(column_line, "tsvfile", append = T)

# Write each raman reading and the corresponding variables as a row in the file
for (i in 1:nrow(df_raman)){
   
    # Sample name as a combination of elapsed time and culture age
    sample_name = paste(df_raman$ElapsedTime[i], df_raman$ProcessID[i], sep='_')
    
    line = paste('#S', sample_name, df_raman$Intensities[i], sep = ', ')
    
    # ${fileout:tsvfile}
    write(line, "tsvfile", append = T)
}
```
