# Joining Raman and HPLC -> Piroutte
## User Guide
This R report generates a file to be used to build a piroutte model that predicts concentrations
of compounds via a raman spectrum. The report matches raman spectra to HPLC samples taken at the
same time to create training data.

### Required Libraries
* Rlabkey
* data.table
* glue

### Adjustable Parameters
The adjustable paramenters are documented throughout the report, but the most essential are directly
below:
    
* *raman_table* corresponds to the name of the dataset containing the raman in this study
* *hplc_table* likewise corresponds to the name of the dataset containing the HPLC in this study
* *subject_id* corresponds to the SubjectID you want to generate a file from.
* *training_names* are the *Field Names* of the HPLC columns that you want to train the model on. This is what
the file will have written to it.
* *table* - set TRUE if you want to see a table displayed of the join before downloading. Set FALSE to hide.

```{r}
raman_table = 'Raman_Transformation'
hplc_table = 'HPLC_Sugars_OrgAcids_BPD'
subject_id = 'B30_35_ATCC32359_WT_1'
training_names = c('Glucose_g_L', 'Xylose_g_L', 'Itaconic_Acid_g_L')
table = TRUE
```
There are some columns that are assumed to exist and that are **required** to have these **field names**
    
1. The raman dataset has columns *ElapsedTime* and *Intensities*
2. The HPLC dataset has a column *CultureAge* that will be aligned with *ElapsedTime*
3. The raman dataset has a Run parameter  *Wavenumbers* that is a look up to a list containing the used wavenumbers in
a column called *WaveNumbers*
    
If any of these parameters change, a replace all should work, but ensure that you have a back up.

```{r, echo=FALSE}
library(Rlabkey)
library(data.table)
library(glue)

# Verbose
options(echo=TRUE)

base_url = labkey.url.base
folder_path = labkey.url.path

```
# Querying 

## Query Commands

The sql strings are first constructed using glue. Any value inside {}
is interpreted as a variable and is replaced in the returend string with the
corresponding value. This allows changing of dataset names.

Ensure that the column names correspond to *Field Names* not *Caption Names.*
    
### Raman Query
Values to get from raman dataset -- only pulls values within 0.15 hours of
an HPLC to narrow down the imported data. Add {raman_table}.FieldName to 
add another column to the table. Joined on SubjectID, so multiple runs
can be stored in the same dataset. Change the desired SubjectID at
the top of the script.

*ElapsedTime* is **required** in order to match the HPLC and Raman.
```{r}

sql_raman = glue("SELECT DISTINCT {raman_table}.SubjectID,
                 {raman_table}.FileName,
                 {raman_table}.ElapsedTime,
                 {raman_table}.Intensities,
				 {raman_table}.Run.ProcessID	
                 FROM {hplc_table}
                 LEFT JOIN {raman_table} ON {raman_table}.SubjectID = {hplc_table}.SubjectID
                 WHERE abs({raman_table}.ElapsedTime-{hplc_table}.SampleID.CultureAge) < 0.15 
                 AND {raman_table}.SubjectID.SubjectID = '{subject_id}'")
```


### HPLC Query
Values to get from the HPLC dataset. Pulls all rows that match the specified SubjectID
at the top of the script. The name of the HPLC dataset can be changed at the top
of the script, just as the raman can. To add another column to the query, add
{hplc_table}.FieldName in the SELECT statement.

*CultureAge* is **required** in order to match the HPLC and Raman.


```{r}
query_targets = glue(paste('{hplc_table}.', training_names, sep='', collapse = ', '))

# Values to get from HPLC table
sql_hplc = glue("SELECT {hplc_table}.SampleID,
                {hplc_table}.SampleID.DateTime,
                {hplc_table}.SampleID.CultureAge,
                {query_targets}
                FROM {hplc_table}
                WHERE {hplc_table}.SubjectID.SubjectID = '{subject_id}'")
```


### Wavenumbers Query
This query gets the wavenumbers corresponding to the SubjectID defined at the top of the script.

The raman dataset is **required** to have a *Run* property called *Wavenumbers* that is a LookUp to 
a list. That list must have the column *WaveNumbers*, which holds the wavenumbers for that configuration.
The list is not required to be in the current study if the raman dataset was imported from an assay, as the
LookUp will automatically reference the orginal list in the assay.
   
   
```{r}
# Getting wavenumbers from raman table -- should only be one when generating piroutte file.
sql_wavenumbers = glue("SELECT DISTINCT
                       {raman_table}.Run.Wavenumbers.WaveNumbers
                       FROM {raman_table}
                       WHERE {raman_table}.SubjectID.SubjectID = '{subject_id}'")
   
```
## Execution of Queries

This section executes the queries defined above and stores them in a dataframe.

The datasets are assumed to be in a *study* and in the **current folder**.
These can be changed by adjusting the *folderPath* parameter (currently set to current folder by LabKey)
and the *schemaName* parameter (for example, from 'study' to 'assay').

The *colNameOpt* parameter tells the query use Field Names in the data frame, as opposed to caption names.

Finally, the data frames returned are copied because LabKey protects them internally, so this is required to use the
data.table format below.
   

```{r}
df_raman = copy(labkey.executeSql(baseUrl = base_url, folderPath = folder_path, schemaName='study',
                                  sql = sql_raman, colNameOpt ='fieldname'))

df_hplc = copy(labkey.executeSql(baseUrl = base_url, folderPath = folder_path, schemaName='study',
                                 sql = sql_hplc, colNameOpt = 'fieldname'))


df_wavenum = copy(labkey.executeSql(baseUrl = base_url, folderPath = folder_path, schemaName='study',
                                    sql = sql_wavenumbers, colNameOpt = 'fieldname'))
                                
                                
```
## Matching Raman ElapsedTime to HPLC CultureDate

Because raman spectra are taken so frequently, we need to find ones that are closest to an HPLC 
sampling time, one spectra for each HPLC. This section accomplishes that, using the narrowed
down raman data that was querried.      

The data.table package is used because of its rolling window functions, which make this very
easy to do - LabKeySQL lacks this implementation, though it is a relatively standard SQL function.
The thought process is to use LabKeySQL to trim the data to ~3 spectrums per HPLC (within 0.15 of an hour)
then use data.table to pick the closest one of those three.

The following code is only commented inline, but is relatively simple. It will **require changing only if**
    
1. The column *ElapsedTime* is changed
2. The column *CultureAge* is changed
3. The column names are changed from field names to caption names (in the executing query section above)

   
```{r}
# Use data table format for access to rolling window function
setDT(df_raman)
setDT(df_hplc)

# Create a copy of the raman ElapsedTime by reference in the raman dataframe to use as a key against HPLC CultureAge, named CultureAge
df_raman[, CultureAge := ElapsedTime]

# Set the keys of both to be CultureAge
setkey(df_raman, CultureAge)
setkey(df_hplc, CultureAge)

# Join tables on a rolling basis. result now contains matched data from both raman and hplc tables.
result = df_raman[df_hplc, roll=Inf]

   
```
## Writing the Piroutte File

This section is responsible for constructing the piroutte file, which is then made avialable to download
for the user by LabKey.

The order is described here:

1. Column names for the target are created by prepending training names with '$' (telling piroutte
they are values to train on) and collapsing them into one string. This is used later in the header.
   
2. The number of intensities are found, which is needed for the dimension line of the input. This is done
by counting the commas in an intensity entry (because they are stored as one continuous string) and adding one.

3. The first line of the file is written, which is the dimension line. It is of the format (# of intensities x # of rows)
   
4. The column line is written, which contains the wavenumbers and the target variable names (*col_names*)

5. The rows of data are looped through and each written. The *FileName* value is used as the sample name, which piroutte requires.
Each row contains the intensities and the training/target values.
   
**Note:** Anytime that you write to a file in LabKey, you must include the comment **# ${fileout:tsvfile** *(closing bracket needed but not shown)* directly above the line with the write call, for **every** write call. *fileout* specifies that you want to make a file
available to download and *tsvfile* is the name of it that you will refer to in the write call. For example:  
   
   # ${fileout:tsvfile  
   write(dimension_line, "tsvfile")
   
See source code for clarification, as LabKey makes the subsitutes in the report.

```{r}
# Create target columns for piruotte by taking training_names, prepending a '$' to each
# and collapsing into a comma separate string
col_names = paste('$', training_names, sep = '', collapse = ', ')

# Get how many intensity entries there are, required for the input dimension size.
# There are stored as a string, so we count the occurances of commas and add one
intensities_len = lengths(regmatches(df_wavenum[1,1], gregexpr(",", df_wavenum[1,1]))) + 1


# First line of file, dimension (number of intensities and rows of data)
dimension_line = paste("#D", paste(intensities_len + length(training_names), nrow(result), sep ='x'), sep =', ')


# EVERYTIME you want to WRITE TO A FILE, the following comment is REQUIRED on the line directly above it
# due to how the LabKey variable subsitutor works

# ${fileout:tsvfile}
write(dimension_line, "tsvfile")

# Write the column names (wavenumber and other variable names)
column_line = paste('#C', df_wavenum[1,1], col_names, sep = ', ')
# ${fileout:tsvfile}
write(column_line, "tsvfile", append = T)

# Write each raman reading and the corresponding variables as a row in the file
for (i in 1:nrow(result)){
    target_values = paste(result[i, ..training_names], collapse = ', ')
   
    # Sample name as a combination of elapsed time and culture age
    sample_name = paste(result$ElapsedTime[i], result$ProcessID[i], sep='_')
    
    line = paste('#S', sample_name, result$Intensities[i], target_values, sep = ', ')
    
    # ${fileout:tsvfile}
    write(line, "tsvfile", append = T)
}

if (table) {
   # ${tsvout:datafile}
   write.table(result[, c("DateTime", "ElapsedTime", "CultureAge")], file = "datafile", sep = "\t",
               qmethod = "double", col.names=NA)
}
```
